---
title: "Intermediate R"
output: html_notebook
---

# Conditionals and Control Flow

## Equality

The most basic form of comparison is equality. Let's briefly recap its syntax.

Notice from the last expression that R is case sensitive: "R" is not equal to "r".

```{r}
# Comparison of logicals
TRUE == FALSE

# Comparison of numerics
-6 * 14 != 17 - 101

# Comparison of character strings
"useR" == "user" 

# Compare a logical with a numeric
TRUE == 1
```

## Greater and less than
You can add an equal sign to express less than or equal to or greater than or equal to, respectively. 

Remember that for string comparison, R determines the greater than relationship based on alphabetical order. Also, keep in mind that TRUE is treated as 1 for arithmetic, and FALSE is treated as 0. Therefore, FALSE < TRUE is TRUE.

```{r}
# Comparison of numerics
-6 * 5 + 2 >= -10 + 1

# Comparison of character strings
"raining" <= "raining dogs"

# Comparison of logicals
TRUE > FALSE
```

## Compare vectors
You are already aware that R is very good with vectors. Without having to change anything about the syntax, R's relational operators also work on vectors.

You want to figure out whether your activity on social media platforms have paid off and decide to look at your results for LinkedIn and Facebook. The sample code in the editor initializes the vectors **linkedin*8 and **facebook**. Each of the vectors contains the number of profile views your LinkedIn and Facebook profiles had over the last seven days.

```{r}
# The linkedin and facebook vectors have already been created for you
linkedin <- c(16, 9, 13, 5, 2, 17, 14)
facebook <- c(17, 7, 5, 16, 8, 13, 14)

# Popular days
linkedin > 15

# Quiet days
linkedin <= 5

# LinkedIn more popular than Facebook
linkedin > facebook
```

## Compare matrices
R's ability to deal with different data structures for comparisons does not stop at vectors. Matrices and relational operators also work together seamlessly!

Instead of in vectors (as in the previous exercise), the LinkedIn and Facebook data is now stored in a matrix called views. The first row contains the LinkedIn information; the second row the Facebook information. The original vectors **facebook** and **linkedin** are still available as well.

```{r}
# The social data has been created for you
linkedin <- c(16, 9, 13, 5, 2, 17, 14)
facebook <- c(17, 7, 5, 16, 8, 13, 14)
views <- matrix(c(linkedin, facebook), nrow = 2, byrow = TRUE)

# When does views equal 13?
views == 13

# When is views less than or equal to 14?
views <= 14
```

## & and |
In this exercise, you'll be working with the **last** variable. This variable equals the last value of the **linkedin** vector that you've worked with previously. The linkedin vector represents the number of LinkedIn views your profile had in the last seven days, remember? 

```{r}
# The linkedin and last variable are already defined for you
linkedin <- c(16, 9, 13, 5, 2, 17, 14)
last <- tail(linkedin, 1)

# Is last under 5 or above 10?
last < 5 | last > 10

# Is last between 15 (exclusive) and 20 (inclusive)?
last > 15 & last <= 20
```

Like relational operators, logical operators work perfectly fine with vectors and matrices. A matrix **views** has been defined; its first and second row correspond to the **linkedin** and **facebook** vectors, respectively. 
```{r}
# linkedin exceeds 10 but facebook below 10
linkedin > 10 & facebook < 10

# When were one or both visited at least 12 times?
linkedin >= 12 | facebook >= 12

# When is views between 11 (exclusive) and 14 (inclusive)?
views > 11 & views <= 14
```

## The if statement
```{r}
# Variables related to your last day of recordings
medium <- "LinkedIn"
num_views <- 14

# Examine the if statement for medium
if (medium == "LinkedIn") {
  print("Showing LinkedIn information")
}

# Write the if statement for num_views
if (num_views >15) {
  print("You are popular!")
}  
```

## Add an else
You can only use an else statement in combination with an if statement. The else statement does not require a condition; its corresponding code is simply run if all of the preceding conditions in the control structure are FALSE. 

It's important that the else keyword comes on the same line as the closing bracket of the if part!

```{r}
# Variables related to your last day of recordings
medium <- "LinkedIn"
num_views <- 14

# Control structure for medium
if (medium == "LinkedIn") {
  print("Showing LinkedIn information")
} else {
  print("Unknown medium") 
}

# Control structure for num_views
if (num_views > 15) {
  print("You're popular!")
} else {
  print("Try to be more visible!")
}
```

## Customize further: else if
The else if statement allows you to further customize your control structure. You can add as many else if statements as you like. Keep in mind that R ignores the remainder of the control structure once a condition has been found that is TRUE and the corresponding expressions have been executed. 

Again, It's important that the else if keywords comes on the same line as the closing bracket of the previous part of the control construct!
```{r}
# Variables related to your last day of recordings
medium <- "LinkedIn"
num_views <- 14

# Control structure for medium
if (medium == "LinkedIn") {
  print("Showing LinkedIn information")
} else if (medium == "Facebook") {
  # Add code to print correct string when condition is TRUE
  print("Showing Facebook information")
} else {
  print("Unknown medium")
}

# Control structure for num_views
if (num_views > 15) {
  print("You're popular!")
} else if (num_views <= 15 & num_views > 10) {
  # Add code to print correct string when condition is TRUE
  print("Your number of views is average")
} else {
  print("Try to be more visible!")
}
```

## Else if 2.0
You can do anything you want inside if-else constructs. You can even put in another set of conditional statements. 

li and fb denote the number of profile views your LinkedIn and Facebook profile had on the last day of recordings. 
```{r}
# Variables related to your last day of recordings
li <- 15
fb <- 9

# Code the control-flow construct
if (li >= 15 & fb >= 15) {
  sms <- 2 * (li + fb)
} else if (li < 10 & fb < 10) {
  sms <- 0.5 * (li + fb)
} else {
  sms <- li + fb
}

# Print the resulting sms to the console
sms
```

# Loops
## Write a while loop
```{r}
# Initialize the speed variable
speed <- 64

# Code the while loop
while (speed > 30) {
  print("Slow down!")
  speed <- speed - 7
}

# Print out the speed variable
speed
```
You simulated the interaction between a driver and a driver's assistant: When the speed was too high, "Slow down!" got printed out to the console, resulting in a decrease of your speed by 7 units.

There are several ways in which you could make your driver's assistant more advanced. For example, the assistant could give you different messages based on your speed or provide you with a current speed at a given moment.

```{r}
# Initialize the speed variable
speed <- 64

# Extend/adapt the while loop
while (speed > 30) {
  print(paste("Your speed is", speed))
  if (speed > 48) {
    print("Slow down big time!")
    speed <- speed - 11
  } else {
    print("Slow down!")
    speed <- speed - 6
  }
}
```

## Stop the while loop: break
There are some very rare situations in which severe speeding is necessary: what if a hurricane is approaching and you have to get away as quickly as possible? You don't want the driver's assistant sending you speeding notifications in that scenario, right?

This seems like a great opportunity to include the break statement in the while loop you've been working on. Remember that the break statement is a control statement. When R encounters it, the while loop is abandoned completely.

```{r}
# Initialize the speed variable
speed <- 88

while (speed > 30) {
  print(paste("Your speed is", speed))
  
  # Break the while loop when speed exceeds 80
  if (speed > 80) {
    break
  }
  
  if (speed > 48) {
    print("Slow down big time!")
    speed <- speed - 11
  } else {
    print("Slow down!")
    speed <- speed - 6
  }
}
```

## Build a while loop from scratch

```{r}
# Initialize i as 1 
i <- 1

# Code the while loop
while (i <= 10) {
  print(3 * i)
  if ( (i * 3) %% 8 == 0 ) {
    break
  }
  i <- i + 1
}
```

## Loop over a vector
```{r}
# The linkedin vector has already been defined for you
linkedin <- c(16, 9, 13, 5, 2, 17, 14)

# Loop version 1
for (li in linkedin) {
  print(li)
}

# Loop version 2
for (i in 1:length(linkedin)) {
  print(linkedin[i])
}
```

## Loop over a list
Looping over a list is just as easy and convenient as looping over a vector.

Suppose you have a list of all sorts of information on New York City: its population size, the names of the boroughs, and whether it is the capital of the United States. 

Notice that you need double square brackets - [[ ]] - to select the list elements in loop version 2.

```{r}
# The nyc list is already specified
nyc <- list(pop = 8405837, 
            boroughs = c("Manhattan", "Bronx", "Brooklyn", "Queens", "Staten Island"), 
            capital = FALSE)

# Loop version 1
for (info in nyc) {
  print(info)
}

# Loop version 2
for (i in 1:length(nyc)) {
  print(nyc[[i]])
}
```

## Loop over a matrix
There's a matrix **ttt**, that represents the status of a tic-tac-toe game. It contains the values "X", "O" and "NA". On row 1 and column 1, there's "O", while on row 3 and column 2 there's "NA".

To solve this exercise, you'll need a for loop inside a for loop, often called a nested loop. Doing this in R is a breeze! 

```{r}
ttt <- matrix(
  c("O", NA, "X", NA, "O", "O", "X", NA, "X"),
  nrow=3,
  ncol=3,
  byrow = TRUE
)
# define the double for loop
for (i in 1:nrow(ttt)) {
  for (j in 1:ncol(ttt)) {
    print(paste("On row", i, "and column", j, "the board contains", ttt[i,j]))
  }
}
```

Let's return to the LinkedIn profile views data, stored in a vector **linkedin**.
```{r}
# The linkedin vector has already been defined for you
linkedin <- c(16, 9, 13, 5, 2, 17, 14)

# Code the for loop with conditionals
for (li in linkedin) {
  if (li > 10) {
    print("You're popular!")
  } else {
    print("Be more visible!")
  }
  print(li)
}
```

## Next, you break it
In this exercise, you will use the break and next statements:

* The break statement abandons the active loop: the remaining code in the loop is skipped and the loop is not iterated over anymore.
* The next statement skips the remainder of the code in the loop, but continues the iteration.
```{r}
# The linkedin vector has already been defined for you
linkedin <- c(16, 9, 13, 5, 2, 17, 14)

# Adapt/extend the for loop
for (li in linkedin) {
  if (li > 10) {
    print("You're popular!")
  } else {
    print("Be more visible!")
  }
  
  # Add if statement with break
  if (li > 16) {
    print("This is ridiculous, I'm outta here!")
    break
  }
  
  # Add if statement with next
  if (li < 5) {
    print("This is too embarrassing!")
    next
  }
  
  print(li)
}
```

## Build a for loop from scratch
**rquote** has been split up into a vector that contains separate letters and has been stored in a vector chars with the strsplit() function.
```{r}
rquote <- "r's internals are irrefutably intriguing"
chars <- strsplit(rquote, split = "")[[1]]

# Initialize rcount
rcount <- 0

# Finish the for loop
for (char in chars) {
  if (char == "r") {
    rcount <- rcount + 1
  }
  if (char == "u") {
    break
  }
}

# Print out rcount
rcount
```

# Functions
## Function documentation
Before even thinking of using an R function, you should clarify which arguments it expects. All the relevant details such as a description, usage, and arguments can be found in the documentation. 
```{r}
# Consult the documentation on the mean() function
help(mean)

# Inspect the arguments of the mean() function
args(mean)
```
The documentation on the mean() function gives us quite some information:

* The mean() function computes the arithmetic mean.
* The most general method takes multiple arguments: x and ....
* The x argument should be a vector containing numeric, logical or time-related information.

Once more, you'll be working with the view counts of your social network profiles for the past 7 days. 

```{r}
# The linkedin and facebook vectors have already been created for you
linkedin <- c(16, 9, 13, 5, 2, 17, 14)
facebook <- c(17, 7, 5, 16, 8, 13, 14)

# Calculate average number of views
avg_li <- mean(x = linkedin)
avg_fb <- mean(facebook)

# Inspect avg_li and avg_fb
avg_li
avg_fb
```

Check the documentation on the mean() function again:

```{r}
?mean
```

The Usage section of the documentation includes two versions of the mean() function. The first usage,
**mean(x, ...)**
is the most general usage of the mean function. The 'Default S3 method', however, is:

**mean(x, trim = 0, na.rm = FALSE, ...)**

The ... is called the ellipsis. It is a way for R to pass arguments along without the function having to name them explicitly. 
```{r}
# The linkedin and facebook vectors have already been created for you
linkedin <- c(16, 9, 13, 5, 2, 17, 14)
facebook <- c(17, 7, 5, 16, 8, 13, 14)

# Calculate the mean of the sum
avg_sum <- mean(linkedin + facebook)

# Calculate the trimmed mean of the sum
avg_sum_trimmed <- mean(linkedin + facebook, trim = 0.2)

# Inspect both new variables
avg_sum
avg_sum_trimmed
```

Let's see what happens if your vectors linkedin and facebook contain missing values (NA).
```{r}
# The linkedin and facebook vectors have already been created for you
linkedin <- c(16, 9, 13, 5, NA, 17, 14)
facebook <- c(17, NA, 5, 16, 8, 13, 14)

# Basic average of linkedin
mean(linkedin)

# Advanced average of linkedin
mean(linkedin, na.rm = TRUE)
```
You already know that R functions return objects that you can then use somewhere else. This makes it easy to use functions inside functions, as you've seen before:

```{r}
speed <- 31
print(paste("Your speed is", speed))
```

Notice that both the print() and paste() functions use the ellipsis - ... - as an argument. 

## Write your own function
Wow, things are getting serious… you're about to write your own function!
```{r}
# Create a function pow_two()
pow_two <- function(x) {
  x ^ 2
}

# Use the function
pow_two(12)

# Create a function sum_abs()
sum_abs <- function(x, y) {
  abs(x) + abs(y)
}

# Use the function
sum_abs(-2, 3)
```
There are situations in which your function does not require an input. Let's say you want to write a function that gives us the random outcome of throwing a fair die:

```{r}
throw_die <- function() {
  number <- sample(1:6, size = 1)
  number
}

throw_die()
```

```{r}
# Define the function hello()
hello <- function() {
  print("Hi there!")
  TRUE
}

# Call the function hello()
hello()
```

You can define default argument values in your own R functions as well. 
```{r}
# Finish the pow_two() function
pow_two <- function(x, print_info = TRUE) {
  y <- x ^ 2
  if (print_info) {
    print(paste(x, "to the power two equals", y))
  }
  return(y)
}

# Some calls of the pow_two() function
pow_two(5)
pow_two(5, FALSE)
pow_two(5, TRUE)
```

## R you functional?
Now that you've acquired some skills in defining functions with different types of arguments and return values, you should try to create more advanced functions. 

Remember our social media example? As a first step, you will be writing a function that can interpret a single value of this vector.
```{r}
linkedin <- c(16, 9, 13, 5, 2, 17, 14)
facebook <- c(17, 7, 5, 16, 8, 13, 14)

# Define the interpret function
interpret <- function(num_views) {
  if (num_views > 15) {
    print("You're popular!")
    return(num_views)
  } else {
    print("Try to be more visible!")
    return(0)
  }
}

# Call the interpret function twice
interpret(linkedin[1])
interpret(facebook[2])
```

Furthermore, your function will return the sum of views on popular days, if asked for. A for loop is ideal for iterating over all the vector elements. The ability to return the sum of views on popular days is something you can code through a function argument with a default value.

```{r}
# The linkedin and facebook vectors have already been created for you
linkedin <- c(16, 9, 13, 5, 2, 17, 14)
facebook <- c(17, 7, 5, 16, 8, 13, 14)

# The interpret() can be used inside interpret_all()
interpret <- function(num_views) {
  if (num_views > 15) {
    print("You're popular!")
    return(num_views)
  } else {
    print("Try to be more visible!")
    return(0)
  }
}

# Define the interpret_all() function
# views: vector with data to interpret
# return_sum: return total number of views on popular days?
interpret_all <- function(views, return_sum = TRUE) {
  count <- 0

  for (v in views) {
    count <- count + interpret(v)
  }

  if (return_sum) {
    return(count)
  } else {
    return(NULL)
  }
}

# Call the interpret_all() function on both linkedin and facebook
interpret_all(linkedin)
interpret_all(facebook)
```

## Load an R Package
There are basically two extremely important functions when it comes down to R packages:

* install.packages(), which as you can expect, installs a given package.
* library() which loads packages, i.e. attaches them to the search list on your R workspace.
* search(), which look at the currently attached packages

To install packages, you need administrator privileges. This means that install.packages() will thus not work in the DataCamp interface. However, almost all CRAN packages are installed on our servers. You can load them with library().

The library() and require() functions are not very picky when it comes down to argument types: both library(rjson) and library("rjson") work perfectly fine for loading a package.

# The apply family
## Use lapply with a built-in R function
To put it generally, lapply takes a vector or list X, and applies the function FUN to each of its members. If FUN requires additional arguments, you pass them after you've specified X and FUN (...). The output of lapply() is a list, the same length as X, where each element is the result of applying FUN on the corresponding element of X.

```{r}
# The vector pioneers has already been created for you
pioneers <- c("GAUSS:1777", "BAYES:1702", "PASCAL:1623", "PEARSON:1857")

# Split names from birth year
split_math <- strsplit(pioneers, split = ":")

# Convert to lowercase strings: split_low
split_low <- lapply(split_math, tolower)

# Take a look at the structure of split_low
str(split_low)
```

## Use lapply with your own function
```{r}
# Code from previous exercise:
pioneers <- c("GAUSS:1777", "BAYES:1702", "PASCAL:1623", "PEARSON:1857")
split <- strsplit(pioneers, split = ":")
split_low <- lapply(split, tolower)

# Write function select_first()
select_first <- function(x) {
  x[1]
}

# Apply select_first() over split_low: names
names <- lapply(split_low, select_first)

# Write function select_second()
select_second <- function(x) {
  x[2]
}

# Apply select_second() over split_low: years
years <- lapply(split_low, select_second)
```

## lapply and anonymous functions
Writing your own functions and then using them inside lapply() is quite an accomplishment! But defining functions to use them only once is kind of overkill, isn't it? That's why you can use so-called anonymous functions in R.

Previously, you learned that functions in R are objects in their own right. This means that they aren't automatically bound to a name. When you create a function, you can use the assignment operator to give the function a name. It's perfectly possible, however, to not give the function a name.
```{r}
# split_low has been created for you
split_low

# Transform: use anonymous function inside lapply
names <- lapply(split_low, function(x) { x[1] })

# Transform: use anonymous function inside lapply
years <- lapply(split_low, function(x) { x[2] })
```
## Use lapply with additional arguments
```{r}
# Definition of split_low
pioneers <- c("GAUSS:1777", "BAYES:1702", "PASCAL:1623", "PEARSON:1857")
split <- strsplit(pioneers, split = ":")
split_low <- lapply(split, tolower)

# Generic select function
select_el <- function(x, index) {
  x[index]
}

# Use lapply() twice on split_low: names and years
names <- lapply(split_low, select_el, index = 1)
years <- lapply(split_low, select_el, index = 2)
```

## How to use sapply
You can use sapply() similar to how you used lapply(). The first argument of sapply() is the list or vector X over which you want to apply a function, FUN. Potential additional arguments to this function are specified afterwards (...).

You'll be working with the variable temp, that contains temperature measurements for 7 days. temp is a list of length 7, where each element is a vector of length 5, representing 5 measurements on a given day. 

```{r}
temp <- list(c(3, 7 ,9 ,6 ,-1),
             c(6, 9, 12, 13, 5),
             c(4,  8,  3, -1, -3),
             c(1, 4, 7, 2, -2),
             c(5, 7, 9, 4, 2),
             c(-3, 5, 8, 9, 4),
             c(3, 6, 9, 4, 1))
str(temp)
# Use lapply() to find each day's minimum temperature
lapply(temp, min)

# Use sapply() to find each day's minimum temperature
sapply(temp, min)

# Use lapply() to find each day's maximum temperature
lapply(temp, max)

# Use sapply() to find each day's maximum temperature
sapply(temp, max)
```
Can you tell the difference between the output of lapply() and sapply()? The former returns a list, while the latter returns a vector that is a simplified version of this list. 

## sapply with your own function
Like lapply(), sapply() allows you to use self-defined functions and apply them over a vector or a list.
```{r}
# Finish function definition of extremes_avg
extremes_avg <- function(x) {
  ( min(x) + max(x) ) / 2
}

# Apply extremes_avg() over temp using sapply()
sapply(temp, extremes_avg)

# Apply extremes_avg() over temp using lapply()
lapply(temp, extremes_avg)
```
You've seen how sapply() simplifies the list that lapply() would return by turning it into a vector. But what if the function you're applying over a list or a vector returns a vector of length greater than 1?
```{r}
# Create a function that returns min and max of a vector: extremes
extremes <- function(x) {
  c(min = min(x), max = max(x))
}

# Apply extremes() over temp with sapply()
sapply(temp, extremes)

# Apply extremes() over temp with lapply()
lapply(temp, extremes)
```

It seems like we've hit the jackpot with sapply(). On all of the examples so far, sapply() was able to nicely simplify the rather bulky output of lapply(). But, as with life, there are things you can't simplify. How does sapply() react?

We create a function, below_zero(), that takes a vector of numerical values and returns a vector that only contains the values that are strictly below zero.
```{r}
# Definition of below_zero()
below_zero <- function(x) {
  return(x[x < 0])
}

# Apply below_zero over temp using sapply(): freezing_s
freezing_s <- sapply(temp, below_zero)

# Apply below_zero over temp using lapply(): freezing_l
freezing_l <- sapply(temp, below_zero)

# Are freezing_s and freezing_l identical?
identical(freezing_s, freezing_l)
```

## sapply with functions that return NULL
You already have some apply tricks under your sleeve, but you're surely hungry for some more, aren't you? Now, you'll see how sapply() reacts when it is used to apply a function that returns NULL over a vector or a list.

A function print_info(), that takes a vector and prints the average of this vector. It uses the cat() function.
```{r}
# Definition of print_info()
print_info <- function(x) {
  cat("The average temperature is", mean(x), "\n")
}

# Apply print_info() over temp using sapply()
sapply(temp, print_info)

# Apply print_info() over temp using lapply()
lapply(temp, print_info)
```
## Use vapply
Before you get your hands dirty with the third and last apply function, let's take a look at its syntax. The function is called vapply(), and it has the following syntax:

```{r}
help(vapply)
```

Over the elements inside X, the function FUN is applied. The FUN.VALUE argument expects a template for the return argument of this function FUN. USE.NAMES is TRUE by default; in this case vapply() tries to generate a named array, if possible.

You'll be working on the temp list again, that contains 7 numerical vectors of length 5. We also coded a function basics() that takes a vector, and returns a named vector of length 3, containing the minimum, mean and maximum value of the vector respectively.

```{r}
# Definition of basics()
basics <- function(x) {
  c(min = min(x), mean = mean(x), max = max(x))
}

# Apply basics() over temp using vapply()
vapply(temp, basics, numeric(3))
```

# Utilities
## Mathematical utilities
Have another look at some useful math functions that R features:

* abs(): Calculate the absolute value.
* sum(): Calculate the sum of all the values in a data structure.
* mean(): Calculate the arithmetic mean.
* round(): Round the values to 0 decimal places by default.

## Data Utilities
R features a bunch of functions to juggle around with data structures::

* seq(): Generate sequences, by specifying the from, to, and by arguments.
* rep(): Replicate elements of vectors and lists.
* sort(): Sort a vector in ascending order. Works on numerics, but also on character strings and logicals.
* rev(): Reverse the elements in a data structures for which reversal is defined.
* str(): Display the structure of any R object.
append(): Merge vectors or lists.
* is.*(): Check for the class of an R object.
* as.*(): Convert an R object from one class to another.
* unlist(): Flatten (possibly embedded) lists to produce a vector.

```{r}
# The linkedin and facebook lists have already been created for you
linkedin <- list(16, 9, 13, 5, 2, 17, 14)
facebook <- list(17, 7, 5, 16, 8, 13, 14)

# Convert linkedin and facebook to a vector: li_vec and fb_vec
li_vec <- unlist(linkedin)
fb_vec <- unlist(facebook)

# Append fb_vec to li_vec: social_vec
social_vec <- append(li_vec, fb_vec)

# Sort social_vec
sort(social_vec, decreasing = TRUE)
```

## Beat Gauss using R
There is a popular story about young Gauss. As a pupil, he had a lazy teacher who wanted to keep the classroom busy by having them add up the numbers 1 to 100. Gauss came up with an answer almost instantaneously, 5050. On the spot, he had developed a formula for calculating the sum of an arithmetic series. There are more general formulas for calculating the sum of an arithmetic series with different starting values and increments. Instead of deriving such a formula, why not use R to calculate the sum of a sequence?
```{r}
# Create first sequence: seq1
seq1 <- seq(1, 500, by = 3)

# Create second sequence: seq2
seq2 <- seq(1200, 900, by = -7)

# Calculate total sum of the sequences
sum(seq1) + sum(seq2)
```

## grepl & grep
In their most basic form, regular expressions can be used to see whether a pattern exists inside a character string or a vector of character strings. For this purpose, you can use:

* grepl(), which returns TRUE when a pattern is found in the corresponding character string.
* grep(), which returns a vector of indices of the character strings that contains the pattern.
Both functions need a pattern and an x argument, where pattern is the regular expression you want to match for, and the x argument is the character vector from which matches should be sought.
```{r}
# The emails vector has already been defined for you
emails <- c("john.doe@ivyleague.edu", "education@world.gov", "dalai.lama@peace.org",
            "invalid.edu", "quant@bigdatacollege.edu", "cookie.monster@sesame.tv")

# Use grepl() to match for "edu"
grepl("edu", emails)

# Use grep() to match for "edu", save result to hits
hits <- grep("edu", emails)

# Subset emails using hits
emails[hits]
```

You can use the caret, ^, and the dollar sign, $ to match the content located in the start and end of a string, respectively. This could take us one step closer to a correct pattern for matching only the ".edu" email addresses from our list of emails. But there's more that can be added to make the pattern more robust:

* @, because a valid email must contain an at-sign.
* .*, which matches any character (.) zero or more times (*). Both the dot and the asterisk are metacharacters. You can use them to match any character between the at-sign and the ".edu" portion of an email address.
* \\.edu$, to match the ".edu" part of the email at the end of the string. The \\ part escapes the dot: it tells R that you want to use the . as an actual character.
```{r}
# The emails vector has already been defined for you
emails <- c("john.doe@ivyleague.edu", "education@world.gov", "dalai.lama@peace.org",
            "invalid.edu", "quant@bigdatacollege.edu", "cookie.monster@sesame.tv")

# Use grepl() to match for .edu addresses more robustly
grepl("@.*\\.edu$", emails)

# Use grep() to match for .edu addresses more robustly, save result to hits
hits <- grep("@.*\\.edu$", emails)
```

## sub & gsub

While grep() and grepl() were used to simply check whether a regular expression could be matched with a character vector, sub() and gsub() take it one step further: you can specify a replacement argument. If inside the character vector x, the regular expression pattern is found, the matching element(s) will be replaced with replacement.sub() only replaces the first match, whereas gsub() replaces all matches.

Suppose that emails vector you've been working with is an excerpt of DataCamp's email database. Why not offer the owners of the .edu email addresses a new email address on the datacamp.edu domain? This could be quite a powerful marketing stunt: Online education is taking over traditional learning institutions! Convert your email and be a part of the new generation!

```{r}
# The emails vector has already been defined for you
emails <- c("john.doe@ivyleague.edu", "education@world.gov", "global@peace.org",
            "invalid.edu", "quant@bigdatacollege.edu", "cookie.monster@sesame.tv")

# Use sub() to convert the email domains to datacamp.edu
sub("@.*\\.edu$", "@datacamp.edu", emails)
```

Regular expressions are a typical concept that you'll learn by doing and by seeing other examples. Before you rack your brains over the regular expression, have a look at the new things that will be used:

* .*: A usual suspect! It can be read as "any character that is matched zero or more times".
* \\s: Match a space. The "s" is normally a character, escaping it (\\) makes it a metacharacter.
* [0-9]+: Match the numbers 0 to 9, at least once (+).
* ([0-9]+): The parentheses are used to make parts of the matching string available to define the replacement. The \\1 in the replacement argument of sub() gets set to the string that is captured by the regular expression [0-9]+.

## Right here, right now
In R, dates are represented by Date objects, while times are represented by POSIXct objects. Under the hood, however, these dates and times are simple numerical values. Date objects store the number of days since the 1st of January in 1970. POSIXct objects on the other hand, store the number of seconds since the 1st of January in 1970.

The 1st of January in 1970 is the common origin for representing times and dates in a wide range of programming languages. There is no particular reason for this; it is a simple convention. Of course, it's also possible to create dates and times before 1970; the corresponding numerical values are simply negative in this case.
```{r}
# Get the current date: today
today <- Sys.Date()

# See what today looks like under the hood
unclass(today)

# Get the current time: now
now <- Sys.time()

# See what now looks like under the hood
unclass(now)
```

## Create and format dates
To create a Date object from a simple character string in R, you can use the as.Date() function. The character string has to obey a format that can be defined using a set of symbols (the examples correspond to 13 January, 1982):

* %Y: 4-digit year (1982)
* %y: 2-digit year (82)
* %m: 2-digit month (01)
* %d: 2-digit day of the month (13)
* %A: weekday (Wednesday)
* %a: abbreviated weekday (Wed)
* %B: month (January)
* %b: abbreviated month (Jan)

The following R commands will all create the same Date object for the 13th day in January of 1982:

```{r}
as.Date("1982-01-13")
as.Date("Jan-13-82", format = "%b-%d-%y")
as.Date("13 January, 1982", format = "%d %B, %Y")
```

Notice that the first line here did not need a format argument, because by default R matches your character string to the formats "%Y-%m-%d" or "%Y/%m/%d".

In addition to creating dates, you can also convert dates to character strings that use a different date notation. For this, you use the format() function. Try the following lines of code:

```{r}
today <- Sys.Date()
format(Sys.Date(), format = "%d %B, %Y")
format(Sys.Date(), format = "Today is a %A!")
```

## Create and format times
Similar to working with dates, you can use as.POSIXct() to convert from a character string to a POSIXct object, and format() to convert from a POSIXct object to a character string. Again, you have a wide variety of symbols:

* %H: hours as a decimal number (00-23)
* %I: hours as a decimal number (01-12)
* %M: minutes as a decimal number
* %S: seconds as a decimal number
* %T: shorthand notation for the typical format %H:%M:%S
* %p: AM/PM indicator

For a full list of conversion symbols, consult the strptime documentation in the console:

```{r}
?strptime
```

```{r}
# Definition of character strings representing times
str1 <- "May 23, '96 hours:23 minutes:01 seconds:45"
str2 <- "2012-3-12 14:23:08"

# Convert the strings to POSIXct objects: time1, time2
time1 <- as.POSIXct(str1, format = "%B %d, '%y hours:%H minutes:%M seconds:%S")
time2 <- as.POSIXct(str2)

# Convert times to formatted strings
format(time1, "%M")
format(time2, "%I:%M %p")
```

## Calculations with Dates
Both Date and POSIXct R objects are represented by simple numerical values under the hood. This makes calculation with time and date objects very straightforward: R performs the calculations using the underlying numerical values, and then converts the result back to human-readable time information again.

You can increment and decrement Date objects, or do actual calculations with them:
```{r}
today <- Sys.Date()
today + 1
today - 1

as.Date("2015-03-12") - as.Date("2015-02-27")
```

## Calculations with Times
Calculations using POSIXct objects are completely analogous to those using Date objects. Try to experiment with this code to increase or decrease POSIXct objects:
```{r}
now <- Sys.time()
now + 3600          # add an hour
now - 3600 * 24     # subtract a day
```

Adding or subtracting time objects is also straightforward:
```{r}
birth <- as.POSIXct("1879-03-14 14:37:23")
death <- as.POSIXct("1955-04-18 03:47:12")
einstein <- death - birth
einstein
```

